---
title: "Delphi to OMOP CDM ETL: Workflow and Design Decisions"
author: "ETLDelphi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Delphi to OMOP CDM ETL: Workflow and Design Decisions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteKeyword{ETL}
  %\VignetteKeyword{OMOP}
  %\VignetteKeyword{CDM}
  %\VignetteKeyword{DuckDB}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

ETLDelphi converts Delphi-2M synthetic health trajectories into OMOP CDM v5.4 tables in DuckDB.

The package is deliberately thin:

- the transformation logic lives in SQL under `inst/sql`
- the R code loads files, substitutes configuration values, executes the SQL in order, and records ETL logs

This vignette describes the current workflow, the main design choices, and the parts of the pipeline that are intentionally configurable.

## Architecture

### Default Schemas

The default configuration uses three schemas:

| Layer | Schema | Purpose |
|------|--------|---------|
| Raw source | `src` | CSVs loaded from the Delphi export |
| Staging | `stg` | typed tables, maps, reject tables, and QC outputs |
| OMOP CDM | `main` | vocabulary tables and final OMOP tables |

You can override the schema names through `default_etl_config()`, but the package is easiest to use when you keep the defaults.

### Core Principles

1. **Deterministic IDs**. The ETL creates stable integer IDs with ordered `ROW_NUMBER()` logic. The same input data yields the same IDs on repeat runs.
2. **SQL-first transforms**. Business rules are encoded in SQL, not in ad hoc R data manipulation.
3. **Observable failures**. Reject tables and QC tables make parse failures, missing joins, and unmapped concepts inspectable after each run.
4. **Configurable type concepts**. The loaders use `config$type_concept_ids`, so type concept assignment can be corrected without editing the SQL files.
5. **Domain correction after loading**. Records can be loaded into a provisional table and then moved into the correct OMOP domain by the domain-routing step.

## Pipeline Structure

`run_etl()` executes every `.sql` file under `inst/sql` in lexicographic order. The pipeline is organized into numbered folders.

### 00_admin

Administrative setup:

- create the `stg` schema
- create ETL log tables
- clear prior `stg` tables so the next run starts from a clean staging layer

This step does not clear `src` or the final CDM tables.

### 10_stage

Source CSVs are staged into typed DuckDB tables.

Common staging behaviors:

- parse dates with multiple `try_strptime()` formats
- cast numerics with `TRY_CAST`
- trim and normalize strings
- write bad rows to `stg.reject_*` tables instead of failing silently

At the end of this phase, the data is still source-shaped, but no longer raw text.

### 20_keys

This phase builds deterministic mapping tables such as:

- `stg.map_person`
- `stg.map_provider`
- `stg.map_visit`
- `stg.map_location`
- `stg.map_care_site`

These bridge source identifiers like `member_id` and `encounter_id` to integer OMOP keys.

### 30_vocab

The vocabulary phase maps source values to OMOP concepts.

Representative mapping tables:

- demographics (`stg.map_gender`, `stg.map_race`)
- visit concepts (`stg.map_visit_concept`)
- drugs and immunizations (`stg.map_drug_order`, `stg.map_immunization`)
- conditions (`stg.map_condition`)
- labs and units (`stg.map_loinc_measurement`, `stg.map_units`, `stg.map_measurement_value`)
- vitals, allergy, and therapy mappings

Two important details:

- source values are preserved even when a standard concept cannot be found
- custom mapping CSVs can override package defaults without changing SQL

### 40_cdm_core

This phase loads the foundational OMOP tables:

- `location`
- `care_site`
- `provider`
- `person`
- `visit_occurrence`
- `note`
- `death`
- `cdm_source`

The load order follows foreign-key dependencies.

### 50_cdm_clinical

This phase loads the event tables:

- `condition_occurrence`
- `drug_exposure`
- `measurement`
- `observation`
- `procedure_occurrence`

It also includes domain routing, which moves rows into the correct OMOP domain when the mapped concept belongs somewhere else.

That routing step is important because Delphi source fields do not always align cleanly with OMOP domains.

### 60_cdm_derived

Derived tables are built from the loaded events. In the current pipeline this includes:

- `observation_period`
- `condition_era`
- `drug_era`
- `dose_era`

### 90_qc

The final phase writes quality tables into `stg`, including:

- row counts
- mapping coverage
- orphan checks
- parse-failure counts
- required-field checks
- domain conformance checks

These tables are designed for post-run review, not just debugging.

## Design Decisions

### Person Loading Requires A Parseable Date Of Birth

`person.year_of_birth` is required in OMOP. ETLDelphi therefore only loads persons with a parseable DOB from the enrollment feed.

If a member appears in clinical events but has no usable DOB, the event may still exist in staging, but that member is excluded from `person` and recorded in `stg.reject_person_missing_dob`.

### Unmapped Standard Concepts Use `0`

When a standard concept cannot be assigned, the ETL writes `0` rather than `NULL` in the standard concept field, and keeps the original source value in the `*_source_value` column.

That makes unmapped rows easy to count and export for cleanup.

### Type Concepts Are Configuration, Not Hardcoded Truth

The SQL loaders contain placeholders such as `{visit_type_concept_id}` and `{drug_type_orders}`. `run_sql_scripts()` replaces those with values from `config$type_concept_ids`.

This is intentional. Type concept choices can vary by vocabulary release and by how strictly you want to satisfy DQD.

### Domain Routing Is Part Of The ETL Contract

The clinical loaders intentionally prioritize source-specific loading first, then correct domain mismatches in `70_domain_routing.sql`.

That means:

- the first load is not always the final OMOP table
- QC review should include `stg.domain_routing_log`

### Custom Mapping Files Are First-Class Inputs

The package supports three external CSV inputs:

- drug-name fallback mappings
- custom concept mappings
- custom NDC overrides

Those files are loaded during the vocabulary phase and are meant to be part of the normal maintenance loop.

## Recommended Operational Workflow

For a clean run:

```{r workflow}
library(ETLDelphi)

con <- init_vocabulary_db(
  db_path = "path/to/delphi.duckdb",
  vocabulary_dir = "path/to/vocabulary_download_v5",
  cdm_schema = "main"
)

init_source_data(con, "path/to/delphi_csvs")

cfg <- default_etl_config()
run_etl(con, config = cfg)
```

For iterative mapping work:

```{r iterative}
# Re-run only the mapping and downstream phases
run_etl(con, from_step = "30_vocab")

# Then review QC outputs
DBI::dbGetQuery(con, "SELECT * FROM stg.qc_mapping_coverage")
DBI::dbGetQuery(con, "SELECT * FROM stg.domain_routing_log")
```

## What This ETL Does Not Try To Hide

This package favors explicitness over pretending the source is cleaner than it is.

It does not:

- infer missing DOBs to force every member into `person`
- silently drop parse failures without a reject trail
- assume every source value has a clean standard concept
- hide domain mismatches instead of logging them

That tradeoff makes the pipeline noisier, but easier to validate and maintain.

## Summary

ETLDelphi is a staged, SQL-driven ETL with:

- deterministic ID assignment
- explicit concept mapping
- configurable type concepts
- domain correction after loading
- built-in QC outputs for review

The fastest way to understand a run is:

1. check `stg.etl_step_log`
2. inspect `stg.qc_mapping_coverage`
3. review `stg.domain_routing_log`
4. inspect the specific `stg.reject_*` tables that show non-zero counts
