---
title: "Delphi to OMOP CDM ETL: Workflow and Design Decisions"
author: "ETLDelphi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{ETL workflow and decisions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteKeyword{ETL}
  %\VignetteKeyword{OMOP}
  %\VignetteKeyword{CDM}
  %\VignetteKeyword{DuckDB}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette documents the **Delphi/Delfi to OMOP CDM v5.4** ETL implemented in **ETLDelphi**: the end-to-end workflow, every major design decision, and how to run and extend it.

The ETL transforms source tables (enrollment, encounter, provider, death, problem, medication_orders, medication_fulfillment, current_medications, immunization, lab_orders, lab_results, vital_sign, allergy, therapy_orders, therapy_actions) into OMOP Common Data Model v5.4 tables in **DuckDB**, using a **3-layer pattern** (raw Ôćĺ staging Ôćĺ CDM), **deterministic integer IDs**, and **explicit concept mapping** with observable failures.

---

## Architecture Overview

### Three-Layer Pattern

| Layer | Schema | Purpose |
|-------|--------|---------|
| **Source** | `src` | Raw tables; all columns varchar as in the Delphi DDL. Pre-loaded by the user. |
| **Staging** | `stg` | Typed and cleaned: parsed dates/numerics, trimmed strings, normalized codes. Reject tables capture parse failures. |
| **CDM** | `cdm` | OMOP CDM v5.4 tables; integer IDs only; concept IDs from vocabulary or 0 when unmapped. |

All transformations are expressed as **SQL scripts** run in a fixed order. The **R package** only orchestrates: it runs those scripts in sequence via DBI + duckdb, substitutes schema names from config, and logs steps (and optionally run metadata) to `stg.etl_run_log` and `stg.etl_step_log`.

### Design Principles

1. **Integer IDs only in CDM** ÔÇö All OMOP tables reference integer primary/foreign keys. Source IDs (e.g. `Member_ID`, `Encounter_ID`) are varchar; the ETL builds deterministic **key maps** (`stg.map_person`, `stg.map_visit`, etc.) that assign stable integer IDs.
2. **Unmapped concepts = 0** ÔÇö Per OMOP convention, missing standard concepts are stored as `0` (not NULL), with the original value preserved in `*_source_value` (and optionally `*_source_concept_id` when a source concept exists).
3. **Failures are observable** ÔÇö Reject tables and QC tables record date/numeric parse failures, missing joins, mapping coverage, and referential integrity so issues can be inspected and fixed.

---

## Workflow: Execution Order

Scripts live under `inst/sql/` in numbered folders. The runner executes **all `.sql` files in lexicographic order** (by folder, then file name). The following is the full sequence and intent of each phase.

### Phase 0 ÔÇö Admin and setup

| File | Purpose |
|------|---------|
| `00_admin/00_create_stg_schema.sql` | Create schema `stg` if not exists. |
| `00_admin/01_create_etl_log_tables.sql` | Create `stg.etl_run_log` (run_id, started_at, ended_at, status, git_sha) and `stg.etl_step_log` (run_id, step_name, file_name, started_at, ended_at, status, rows_affected, notes). |
| `00_admin/02_clear_stg_tables.sql` | Drop all `stg.*` tables (reject, maps, staged, QC) so the ETL can be re-run. Does **not** drop `cdm.*` or `src.*`. |

### Phase 1 ÔÇö Staging (typing and normalization)

Source columns are varchar. Each staging script parses dates (with `COALESCE(try_strptime(..., fmt1), try_strptime(..., fmt2), ...)`), casts numerics with `try_cast`, and trims/normalizes strings. Rows that fail parsing (e.g. non-null date string but no parseable date) are written to a corresponding **reject** table.

| File | Staging table | Reject table | Notes |
|------|----------------|--------------|--------|
| `10_stage/10_stage_enrollment.sql` | `stg.enrollment` | `stg.reject_enrollment_dates` | DOB parsed; gender/race/address trimmed and lowercased. |
| `11_stage_provider.sql` | `stg.provider` | `stg.reject_provider` | DOB, sex, specialty cleaned. |
| `12_stage_encounter.sql` | `stg.encounter` | `stg.reject_encounter` | Encounter_DateTime Ôćĺ encounter_date/datetime; appt_type, clinic_type, SOAP_Note. |
| `13_stage_death.sql` | `stg.death` | `stg.reject_death` | DOD Ôćĺ death_date, death_datetime. |
| `14_stage_problem.sql` | `stg.problem` | `stg.reject_problem` | Onset_Date, Resolution_Date; Problem_Code/Type normalized. |
| `15_stage_med_orders.sql` | `stg.medication_orders` | `stg.reject_med_orders` | Order_Date, Last_Filled_Date; Dose, Qty_Ordered, Refills via try_cast; Drug_NDC normalized (strip dashes). |
| `16_stage_med_fulfillment.sql` | `stg.medication_fulfillment` | `stg.reject_med_fulfillment` | Dispense_Date; Dispense_Qty, Days_Of_Supply, Fill_No. |
| `17_stage_current_meds.sql` | `stg.current_medications` | `stg.reject_current_meds` | Last_Filled_Date; Refills, Days_Of_Supply. |
| `18_stage_immunization.sql` | `stg.immunization` | `stg.reject_immunization` | Vaccination_Date; Dose; Vaccine_CVX trimmed. |
| `19_stage_lab_orders.sql` | `stg.lab_orders` | `stg.reject_lab_orders` | Order_Date; Test_LOINC normalized. |
| `20_stage_lab_results.sql` | `stg.lab_results` | `stg.reject_lab_results` | Date_Collected, Date_Resulted; Numeric_Result; LOINC/Units. |
| `21_stage_vital_sign.sql` | `stg.vital_sign` | `stg.reject_vital_sign` | Encounter_Date; Height, Weight, SBP, DBP, Pulse, Respiration, Temperature. |
| `22_stage_allergy.sql` | `stg.allergy` | `stg.reject_allergy` | Onset_Date; Drug_Code, Allergen, etc. normalized. |
| `23_stage_therapy_orders.sql` | `stg.therapy_orders` | ÔÇö | Code, Vocabulary; no dates in source. |
| `24_stage_therapy_actions.sql` | `stg.therapy_actions` | ÔÇö | Code, Vocabulary, Result, Target_Area. |

### Phase 2 ÔÇö Keys (deterministic integer ID maps)

Maps assign **stable integer IDs** to source keys using `ROW_NUMBER() OVER (ORDER BY <stable_key>)`. No randomness; same source data yields same IDs across runs.

| File | Map table | Source of keys |
|------|-----------|----------------|
| `20_keys/30_map_person.sql` | `stg.map_person(member_id, person_id)` | Distinct `member_id` from union of all tables that contain Member_ID (enrollment, encounter, death, problem, medication_orders, immunization, lab_results, vital_sign, allergy, therapy_orders, therapy_actions, current_medications). Ensures persons who appear only outside enrollment still get an ID. |
| `31_map_provider.sql` | `stg.map_provider(provider_id_source, provider_id)` | Distinct Provider_ID from provider, encounter, medication_orders, immunization, lab_results, problem. |
| `32_map_visit.sql` | `stg.map_visit(encounter_id_source, visit_occurrence_id)` | Distinct Encounter_ID from `stg.encounter`. |
| `33_map_location.sql` | `stg.map_location(location_key, location_id)` | Distinct concatenation of address_1, address_2, city, state, zip from `stg.enrollment`. |
| `34_map_care_site.sql` | `stg.map_care_site(care_site_key, care_site_id)` | Distinct Clinic_ID from `stg.encounter` (optional). |

### Phase 3 ÔÇö Vocabulary and concept mapping

Mapping tables/views resolve source values to OMOP concept IDs. Unmapped Ôćĺ `0`; source values are always preserved in the CDM `*_source_value` fields.

| File | Map table | Strategy |
|------|-----------|----------|
| `30_vocab/40_map_gender.sql` | `stg.map_gender` | Lookup: female/fÔćĺ8532, male/mÔćĺ8507, unknown/otherÔćĺ8551; enrollment-specific values inserted with 0. |
| `41_map_race.sql` | `stg.map_race` | Lookup: white, black, asian, etc. to standard race concept_ids; rest Ôćĺ 0. |
| `42_map_visit_concept.sql` | `stg.map_visit_concept` | (appt_type, clinic_type) Ôćĺ visit_concept_id; default 44813942 (Outpatient). |
| `43_map_drug.sql` | `stg.map_drug_order` | NDC Ôćĺ `cdm.concept` (vocabulary_id='NDC') Ôćĺ concept_relationship 'Maps to' Ôćĺ standard RxNorm; drug_name-only rows added with concept_id 0. |
| `44_map_loinc_measurement.sql` | `stg.map_loinc_measurement` | Test_LOINC Ôćĺ `cdm.concept` (vocabulary_id='LOINC'); unmapped Ôćĺ 0. |
| `45_map_units.sql` | `stg.map_units` | Unit strings (mg, cm, kg, etc.) Ôćĺ unit_concept_id; distinct units from staging added as 0 if not in lookup. |
| `46_map_condition.sql` | `stg.map_condition` | Problem_Code Ôćĺ concept (ICD10/ICD9/SNOMED/Medcin) Ôćĺ Maps to standard condition; unmapped Ôćĺ 0. |
| `47_map_vitals.sql` | `stg.map_vitals` | Hardcoded: vital name (Height, Weight, SystolicBP, ÔÇŽ) Ôćĺ measurement_concept_id and default unit (LOINC/OMOP placeholders). |
| `48_map_allergy.sql` | `stg.map_allergy` | v1: observation_concept_id = 0; source_value = Allergen or Drug_Code. |
| `49_map_therapy.sql` | `stg.map_therapy` | Code + Vocabulary Ôćĺ concept (CPT4, HCPCS, SNOMED, ICD10PCS) Ôćĺ Maps to standard procedure; else 0. |

### Phase 4 ÔÇö CDM core tables

Load order respects foreign keys: location and care_site first, then provider, person, visit, note, death.

| File | CDM table | Source and decisions |
|------|-----------|----------------------|
| `40_cdm_core/50_load_location.sql` | `cdm.location` | From `stg.map_location`; address fields from first matching enrollment row per location_key. Insert only if location_id not already present. |
| `51_load_care_site.sql` | `cdm.care_site` | From `stg.map_care_site`; optional. |
| `52_load_provider.sql` | `cdm.provider` | From map_provider + stg.provider; specialty_concept_id null in v1. |
| `53_load_person.sql` | `cdm.person` | From map_person + stg.enrollment. **Only persons with parseable DOB in enrollment** are inserted (year_of_birth NOT NULL). Others logged in `stg.reject_person_missing_dob`. Demographics from enrollment; location_id from map_location. |
| `54_load_visit_occurrence.sql` | `cdm.visit_occurrence` | From stg.encounter + map_visit, map_person, map_visit_concept, map_provider, map_care_site. visit_type_concept_id = 44818517 (EHR). Visit end date = start when no end time. Encounters with member_id not in map_person Ôćĺ `stg.reject_visit_missing_person`. |
| `55_load_note.sql` | `cdm.note` | SOAP_Note from encounter; note_id = 900000000 + row_number (deterministic). note_type/class/encoding/language from config. Reject table for notes missing person/visit. |
| `56_load_death.sql` | `cdm.death` | From stg.death + map_person. death_type_concept_id from config. Reject rows with no matching person Ôćĺ `stg.reject_death_load`. |
| `57_load_cdm_source.sql` | `cdm.cdm_source` | Single row: source name, abbreviation, holder, description, ETL reference (from package vignettes), source/cdm release dates (CURRENT_DATE), cdm_version 5.4, cdm_version_concept_id from vocabulary, vocabulary_version. Insert only if cdm_source is empty. |

### Phase 5 ÔÇö CDM clinical tables

| File | CDM table | Source and decisions |
|------|-----------|----------------------|
| `50_cdm_clinical/60_load_condition_occurrence.sql` | `cdm.condition_occurrence` | stg.problem; condition_occurrence_id deterministic; condition_start_date = onset_date, fallback to encounter_date when onset missing; condition_concept_id from map_condition (0 if unmapped). Reject: missing person. |
| `61_load_drug_exposure_orders.sql` | `cdm.drug_exposure` | stg.medication_orders; drug_type 38000177 (EHR order). drug_exposure_id in 700M range. End date = order_date when no days_supply. |
| `62_load_drug_exposure_fulfillment.sql` | `cdm.drug_exposure` | stg.medication_fulfillment joined to medication_orders for member_id and drug; drug_type 38000230 (dispensed). End date = dispense_date + days_supply - 1 when present. Rows with Order_ID not in orders Ôćĺ `stg.reject_fulfillment_no_order`. |
| `63_load_drug_exposure_immunization.sql` | `cdm.drug_exposure` | stg.immunization; drug_type 38000280 (immunization). drug_concept_id 0 in v1 (CVX mapping optional). |
| `64_load_measurement_vitals.sql` | `cdm.measurement` | stg.vital_sign unpivoted (one row per non-null Height, Weight, SBP, DBP, Pulse, Respiration, Temperature). measurement_concept_id from map_vitals; unit from map_units or map_vitals default. |
| `65_load_measurement_labs.sql` | `cdm.measurement` | stg.lab_results; measurement_concept_id from map_loinc_measurement (0 if unmapped). measurement_date = date_resulted else date_collected. Reject: missing person or measurement_date. |
| `66_load_observation_allergy.sql` | `cdm.observation` | stg.allergy; observation_concept_id 0 in v1; observation_type 32859. observation_source_value = Allergen or Drug_Code. Reject when onset_date missing or person missing. |
| `67_load_procedure_therapy.sql` | `cdm.procedure_occurrence` | stg.therapy_orders + therapy_actions; procedure_date from encounter via Encounter_ID. procedure_concept_id from map_therapy (0 if unmapped). Reject: missing person or missing encounter_date. |

### Phase 6 ÔÇö Derived CDM

| File | CDM table | Logic |
|------|-----------|--------|
| `60_cdm_derived/80_load_observation_period.sql` | `cdm.observation_period` | Per person: min/max of all event dates from visit_occurrence, condition_occurrence, drug_exposure, measurement, observation, procedure_occurrence, death. observation_period_id in 900M range. period_type_concept_id 32821. Insert only for persons not already having an observation_period. |

### Phase 7 ÔÇö QC and observability

| File | QC table | Content |
|------|-----------|---------|
| `90_qc/90_row_counts.sql` | `stg.qc_row_counts` | Row counts for key stg.* and cdm.* tables (run_id from etl_run_log). |
| `91_mapping_coverage.sql` | `stg.qc_mapping_coverage` | Per clinical table: total rows, unmapped count, unmapped % (*_concept_id = 0). |
| `92_referential_integrity.sql` | `stg.qc_orphans` | Orphan checks: visit_occurrence.person_id not in person; condition/drug/measurement person_id not in person; visit_occurrence_id not in visit_occurrence when not null. |
| `93_parse_failures.sql` | `stg.qc_parse_failures` | Count of rows in each `stg.reject_*` table. |
| `94_required_fields.sql` | `stg.qc_required_fields` | Violations of NOT NULL (person.year_of_birth, visit dates, condition_start_date, etc.). Runner can be extended to fail if any violation_count > 0. |

---

## Key Design Decisions

### Deterministic integer IDs

- **person_id, provider_id, visit_occurrence_id, location_id, care_site_id** come from `stg.map_*` tables built with `ROW_NUMBER() OVER (ORDER BY <stable_source_key>)`. Same source data Ôçĺ same IDs on every run.
- **Clinical table IDs** (condition_occurrence_id, drug_exposure_id, measurement_id, observation_id, procedure_occurrence_id) use fixed numeric ranges (e.g. 800M, 700M, 500M, 400M, 300M, 200M, 100M) plus row_number so they are deterministic and non-overlapping.

### Unmapped concepts Ôćĺ 0

- OMOP requires concept_id columns to be NOT NULL where applicable; "no mapping" is represented as **0** (config: `concept_id_unknown`), not NULL.
- Original values are always stored in `*_source_value` (and `*_source_concept_id` when a source concept exists but no standard mapping).

### Person and DOB

- **year_of_birth** in `cdm.person` is NOT NULL. The ETL only inserts persons who have a **parseable DOB in `stg.enrollment`**. Members who appear only in other tables or have null/unparseable DOB are **not** inserted into person and are logged in `stg.reject_person_missing_dob`. No sentinel year or inference from encounter.Patient_DOB in v1.

### Medication orders vs. fulfillment

- **Both** medication_orders and medication_fulfillment are loaded into `cdm.drug_exposure` with different **drug_type_concept_id** (38000177 order, 38000230 dispensed).
- Fulfillment rows are joined to medication_orders on Order_ID to get member_id and drug fields; if Order_ID is missing in orders, the row is logged in `stg.reject_fulfillment_no_order` and not loaded.
- Config flag **prefer_fulfillment** is documented; if set true, the ETL could exclude order rows that have a matching fulfillment (logic can be added in 61/62).

### Visit and encounter

- **Encounter_ID** is the backbone for visit linking. visit_occurrence_id = map_visit(encounter_id). When Encounter_ID is missing on a clinical row, visit_occurrence_id is left NULL but person_id and date are still loaded.
- **Visit end date**: When the source has no end time, visit_end_date = visit_start_date (same-day encounter).

### Reject tables

- Every staging script that parses dates or numerics writes failing rows to a **reject** table (e.g. non-null DOB that did not parse Ôćĺ `stg.reject_enrollment_dates`). Load scripts add rejects for missing person (e.g. `reject_visit_missing_person`, `reject_condition_missing_person`). QC script 93 summarizes counts per reject table.

### Allergy and therapy

- **Allergy**: v1 uses observation_concept_id = 0; observation_source_value holds Allergen or Drug_Code. Optional future: map Drug_Code/Drug_Vocab to drug concepts.
- **Therapy**: procedure_date is taken from encounter via Encounter_ID. Rows with no encounter or no encounter_date are rejected.

---

## Configuration

Configuration is read from **`inst/extdata/config.yml`** (or a path you pass). Key sections:

- **schemas**: `src`, `stg`, `cdm` ÔÇö used for schema substitution in SQL (`@cdmDatabaseSchema`, `{cdm}`, `{stg}`, `{src}`).
- **concept_id_unknown**: `0`.
- **date_formats**: list of strptime formats (used conceptually; SQL scripts use explicit `try_strptime` sequences).
- **type_concept_ids**: visit_type (44818517), default_visit (44813942), condition (32818), drug orders (38000177), dispensed (38000230), immunization (38000280), measurement vitals (32817), labs (32827), observation allergy (32859), note type/class (44813942, 44814639), language (4180186), encoding (44815386), procedure (38000268), period (32821), death (32817).
- **prefer_fulfillment**: when true (future), exclude medication_orders that have a matching fulfillment.

You can override config by passing a list to `run_etl(con, config = my_list)` or a path with `run_etl(con, config_path = "path/to/config.yml")`.

---

## Running the ETL

### Full run

```{r run-full}
library(DBI)
library(duckdb)
library(ETLDelphi)

con <- dbConnect(duckdb::duckdb(), "path/to/your.duckdb")
run_etl(con)
dbDisconnect(con, shutdown = TRUE)
```

### Dry run (list scripts only)

```{r dry-run}
run_etl(con, dry_run = TRUE)
```

### From / to step

Run only a subset of the pipeline (e.g. from CDM core through clinical, or only QC):

```{r from-to}
# From admin through staging
run_etl(con, from_step = "00_admin", to_step = "02_clear")

# From CDM core through end (assumes stg already populated)
run_etl(con, from_step = "40_cdm_core")

# Only QC
run_etl(con, from_step = "90_qc", to_step = "94_required_fields")
```

### Custom config or SQL directory

```{r custom}
run_etl(con, config_path = "/path/to/config.yml")
run_etl(con, sql_dir = "/path/to/sql")
```

---

## QC and Observability

After a run, inspect:

- **`stg.qc_row_counts`** ÔÇö row counts per table.
- **`stg.qc_mapping_coverage`** ÔÇö unmapped counts and percentages per clinical table.
- **`stg.qc_orphans`** ÔÇö referential integrity (orphan person_id, visit_occurrence_id).
- **`stg.qc_parse_failures`** ÔÇö counts per reject table.
- **`stg.qc_required_fields`** ÔÇö NOT NULL violations.

Reject tables (e.g. `stg.reject_enrollment_dates`, `stg.reject_visit_missing_person`) hold the actual failing rows for review. The runner can be extended to **abort** if `stg.qc_required_fields` has any `violation_count > 0`.

---

## Updating Mappings and Limitations

- **Gender/race**: Edit `40_map_gender.sql` and `41_map_race.sql`; add (source_value, concept_id) or rely on the INSERT that adds enrollment values with 0.
- **Visit concept**: Adjust rules in `42_map_visit_concept.sql` (appt_type, clinic_type Ôćĺ visit_concept_id).
- **Drug**: NDC mapping uses `cdm.concept` and `cdm.concept_relationship`; extend for drug name if needed.
- **Condition**: `46_map_condition.sql` uses Problem_Code and vocabulary (ICD10, ICD9, SNOMED, Medcin).
- **Units**: Add unit strings in `45_map_units.sql`.
- **Vitals**: `47_map_vitals.sql` uses placeholder concept IDs; replace with your vocabularyÔÇÖs measurement and unit concepts.

**Known limitations**: Person load requires DOB in enrollment; SOAP_Note may be truncated (varchar(255) in source); allergy observation concept is 0 in v1; observation_period is derived from event dates (no death truncation rule applied in v1). The **prefer_fulfillment** flag is not yet implemented in the load scripts.

---

## Summary

The ETL is a **SQL-first, ordered pipeline** with a clear **src Ôćĺ stg Ôćĺ cdm** flow, **deterministic integer IDs**, **explicit concept mapping** (unmapped = 0), and **reject + QC tables** for observability. The R package **orchestrates** execution and logging; all business logic lives in the SQL scripts under `inst/sql/`, so changes to rules or mappings are made there and documented in this vignette.
