---
title: "How to Run ETLDelphi"
author: "ETLDelphi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{How to Run ETLDelphi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The recommended ETLDelphi workflow has three steps:

1. Create a fresh DuckDB database and load the OMOP vocabulary with `init_vocabulary_db()`.
2. Load the Delphi CSV files into the `src` schema with `init_source_data()`.
3. Run the core SQL ETL with `run_etl()`.

By default, the package uses these schemas:

- `src` for raw Delphi source tables
- `stg` for typed staging, mapping, reject, and QC tables
- `main` for OMOP CDM and vocabulary tables

Those defaults come from `default_etl_config()`. You only need to override them when your database layout is different.

## What You Need

Before running the ETL, you need:

- a directory containing OMOP vocabulary files such as `CONCEPT.csv`, `CONCEPT_RELATIONSHIP.csv`, and `VOCABULARY.csv`
- a directory containing the Delphi source CSV files such as `enrollment.csv`, `encounter.csv`, `lab_results.csv`, and `medication_orders.csv`
- the `DBI`, `duckdb`, and `ETLDelphi` packages installed

The package creates the CDM tables and vocabulary tables for you. You do not need to pre-create the OMOP schema manually when you use `init_vocabulary_db()`.

## End-to-End Run

This is the shortest reliable way to run the ETL from scratch.

```{r end-to-end}
library(DBI)
library(duckdb)
library(ETLDelphi)

db_path <- "path/to/delphi.duckdb"
vocabulary_dir <- "path/to/vocabulary_download_v5"
source_dir <- "path/to/delphi_csvs"

con <- init_vocabulary_db(
  db_path = db_path,
  vocabulary_dir = vocabulary_dir,
  cdm_schema = "main"
)

init_source_data(con, source_dir)

cfg <- default_etl_config()
run_etl(con, config = cfg)

DBI::dbDisconnect(con, shutdown = TRUE)
```

What this does:

- creates the DuckDB file if it does not exist
- creates the OMOP CDM 5.4 tables in `main`
- loads the vocabulary into `main`
- recreates `src` and loads the Delphi CSVs
- runs the ordered SQL pipeline under `inst/sql`

## Common Run Modes

### Dry Run

Use this to see which SQL files would execute without changing the database.

```{r dry-run}
con <- DBI::dbConnect(duckdb::duckdb(), db_path)
run_etl(con, dry_run = TRUE)
DBI::dbDisconnect(con, shutdown = TRUE)
```

### Re-run From a Later Step

If the source and staging data are already loaded, you can restart from a later phase.

```{r rerun-subset}
con <- DBI::dbConnect(duckdb::duckdb(), db_path)

# Rebuild vocabulary-driven maps and everything after them
run_etl(con, from_step = "30_vocab")

# Run only the QC phase
run_etl(con, from_step = "90_qc")

DBI::dbDisconnect(con, shutdown = TRUE)
```

### Stop At A Specific Step

This is useful when debugging a specific phase.

```{r to-step}
con <- DBI::dbConnect(duckdb::duckdb(), db_path)

# Stop after visit_occurrence is loaded
run_etl(con, to_step = "54_load_visit_occurrence")

DBI::dbDisconnect(con, shutdown = TRUE)
```

## Configuration

`run_etl()` uses `default_etl_config()` when you do not pass a config. That object includes:

- schema names
- the sentinel concept ID for unmapped concepts (`0L`)
- accepted date formats for staging
- standard OMOP type concept IDs used by the loaders
- optional paths for custom mapping CSVs

To override a setting, start from the defaults and edit only what you need.

```{r config}
cfg <- default_etl_config()

# Example: use a different CDM schema
cfg$schemas$cdm <- "omop"

# Example: point to project-specific mapping overrides
cfg$custom_mapping_path <- "inst/extdata/custom_concept_mapping.csv"
cfg$custom_ndc_mapping_path <- "inst/extdata/custom_ndc_mapping.csv"

con <- DBI::dbConnect(duckdb::duckdb(), db_path)
run_etl(con, config = cfg)
DBI::dbDisconnect(con, shutdown = TRUE)
```

If you change `cfg$schemas$cdm`, pass the same schema name to `init_vocabulary_db(cdm_schema = ...)` when you create the database.

## What To Check After The Run

The ETL writes useful diagnostics into `stg`:

- `stg.etl_run_log` and `stg.etl_step_log` show which SQL files ran
- `stg.qc_row_counts` summarizes table populations
- `stg.qc_mapping_coverage` shows how many rows still use concept ID `0`
- `stg.qc_parse_failures` summarizes reject-table counts
- `stg.qc_required_fields` highlights required-field violations
- `stg.domain_routing_log` records rows that were moved into a different OMOP domain

Example queries:

```{r post-run-checks}
con <- DBI::dbConnect(duckdb::duckdb(), db_path)

DBI::dbGetQuery(con, "SELECT * FROM stg.qc_row_counts ORDER BY table_name")
DBI::dbGetQuery(con, "SELECT * FROM stg.qc_mapping_coverage ORDER BY cdm_table")
DBI::dbGetQuery(con, "SELECT * FROM stg.qc_parse_failures ORDER BY reject_table")

DBI::dbDisconnect(con, shutdown = TRUE)
```

## Optional: Genomic Extension

The genomic simulation and genomic CDM extension are separate from the core ETL. Run them only after the core ETL has completed successfully.

```{r genomic}
library(CDMConnector)

con <- DBI::dbConnect(duckdb::duckdb(), db_path)

cdm <- cdmFromCon(con, cdmSchema = "main", writeSchema = "main")
cdm <- simulateGenomicData(cdm, overwrite = TRUE)

run_genomic_etl(con, config = default_etl_config())

cdmDisconnect(cdm)
DBI::dbDisconnect(con, shutdown = TRUE)
```

## Summary

For most users, the only functions you need are:

- `init_vocabulary_db()`
- `init_source_data()`
- `run_etl()`

Start with `default_etl_config()`, override only the settings you need, and check the `stg.qc_*` tables after every run.
