---
title: "Oncology Data Diagnostic"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Oncology Data Diagnostic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# Overview

This diagnostic evaluates a CDM dataset for suitability in oncology research. We
examine the presence and depth of cancer-related data across four clinical
domains: **conditions**, **drugs**, **procedures**, and **measurements**. The
report also looks for evidence of cancer staging, disease progression, and tumor
markers -- data elements critical for most oncology study designs.

All concept IDs are hardcoded OMOP standard concepts verified against the OMOP
Standardized Vocabulary.

# Setup

Connect to the CDM database and load required libraries.

```{r connect}
library(DBI)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(CDMConnector)

con <- dbConnect(duckdb::duckdb(), eunomiaDir("delphi-100k"))
cdm <- cdmFromCon(con, cdmSchema = "main", writeSchema = "main", achillesSchema = "main")
```

```{r theme, include=FALSE}
theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold"),
      axis.text.y = element_text(size = 10)
    )
)

fmt <- function(x) format(x, big.mark = ",", scientific = FALSE)
```

# 1. Overall Cancer Prevalence

We start by counting how many persons and condition records are associated with
any malignant neoplasm. The parent concept **443392** (*Malignant neoplastic
disease*) captures all cancers via the SNOMED hierarchy and `concept_ancestor`.

```{r cancer-prevalence}
# 443392 = Malignant neoplastic disease (SNOMED 363346000)
cancer_descendants <- cdm$concept_ancestor %>%
  filter(ancestor_concept_id == 443392L) %>%
  select(descendant_concept_id)

cancer_conditions <- cdm$condition_occurrence %>%
  inner_join(cancer_descendants,
             by = c("condition_concept_id" = "descendant_concept_id"))

cancer_summary <- cancer_conditions %>%
  summarise(
    cancer_persons = n_distinct(person_id),
    cancer_records = n()
  ) %>%
  collect()

total_persons <- cdm$person %>% tally() %>% pull(n)

cancer_summary <- cancer_summary %>%
  mutate(
    total_persons = total_persons,
    prevalence_pct = round(cancer_persons / total_persons * 100, 2)
  )
```

```{r cancer-prevalence-table}
cancer_summary %>%
  transmute(
    `Total Persons in CDM`   = fmt(total_persons),
    `Persons with Cancer`    = fmt(cancer_persons),
    `Cancer Prevalence (%)`  = prevalence_pct,
    `Cancer Condition Records`= fmt(cancer_records)
  ) %>%
  knitr::kable(align = "r")
```

# 2. Top Cancer Subtypes

The table below identifies the most common cancer subtypes by person count. We
query condition_occurrence against concept_ancestor descendants of the
malignant neoplasm hierarchy and resolve concept names.

```{r cancer-subtypes}
cancer_by_type <- cancer_conditions %>%
  group_by(condition_concept_id) %>%
  summarise(
    persons = n_distinct(person_id),
    records = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(persons)) %>%
  head(20) %>%
  left_join(
    cdm$concept %>% select(concept_id, concept_name),
    by = c("condition_concept_id" = "concept_id")
  ) %>%
  collect()
```

```{r cancer-subtypes-plot, fig.height=7}
cancer_by_type %>%
  mutate(label = paste0(concept_name, " [", condition_concept_id, "]")) %>%
  mutate(label = reorder(label, persons)) %>%
  ggplot(aes(x = persons, y = label)) +
  geom_col(fill = "#2c7fb8") +
  geom_text(aes(label = fmt(persons)), hjust = -0.1, size = 3.2) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.25)),
                     labels = scales::comma) +
  labs(
    title = "Top 20 Cancer Subtypes by Person Count",
    x = "Persons", y = NULL
  )
```

We also check against a curated list of common cancer sites to ensure the
hierarchy-based counts are plausible. These are the top-level SNOMED concepts
for each major cancer type.

```{r cancer-sites}
cancer_sites <- tribble(
  ~concept_id, ~cancer_site,
  4112853L,    "Breast",
  443388L,     "Lung",
  4180790L,    "Colon",
  4163261L,    "Prostate",
  141232L,     "Melanoma (skin)",
  4181351L,    "Ovary",
  197508L,     "Bladder",
  4180793L,    "Pancreas",
  432571L,     "Lymphoma",
  317510L,     "Leukemia"
)

site_counts <- cancer_sites %>%
  rowwise() %>%
  mutate(
    descendants = list(
      cdm$concept_ancestor %>%
        filter(ancestor_concept_id == concept_id) %>%
        pull(descendant_concept_id)
    )
  ) %>%
  ungroup() %>%
  mutate(
    stats = purrr::map2(concept_id, descendants, function(cid, descs) {
      ids <- unique(c(cid, descs))
      cdm$condition_occurrence %>%
        filter(condition_concept_id %in% ids) %>%
        summarise(persons = n_distinct(person_id), records = n()) %>%
        collect()
    })
  ) %>%
  tidyr::unnest(stats) %>%
  select(cancer_site, concept_id, persons, records) %>%
  arrange(desc(persons))
```

```{r cancer-sites-table}
site_counts %>%
  mutate(across(c(persons, records), fmt)) %>%
  knitr::kable(
    col.names = c("Cancer Site", "Ancestor Concept ID", "Persons", "Records"),
    align = c("l", "r", "r", "r")
  )
```

```{r cancer-sites-plot}
site_counts %>%
  filter(persons > 0) %>%
  mutate(cancer_site = reorder(cancer_site, persons)) %>%
  ggplot(aes(x = persons, y = cancer_site)) +
  geom_col(fill = "#d95f0e") +
  geom_text(aes(label = fmt(persons)), hjust = -0.1, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.25)),
                     labels = scales::comma) +
  labs(
    title = "Cancer Persons by Site",
    x = "Persons", y = NULL
  )
```

# 3. Chemotherapy & Antineoplastic Drugs

This section examines drug exposures for common chemotherapy agents. We check
both traditional cytotoxic drugs and newer targeted/immunotherapy agents. All
concept IDs below are RxNorm **Ingredient**-level standard concepts; we use
`concept_ancestor` to capture all formulations and strengths.

```{r chemo-drugs}
chemo_drugs <- tribble(
  ~concept_id, ~drug_name,         ~drug_class,
  1310317L,    "Cyclophosphamide",  "Cytotoxic",
  1338512L,    "Doxorubicin",       "Cytotoxic",
  955632L,     "Fluorouracil",      "Cytotoxic",
  1397599L,    "Cisplatin",         "Cytotoxic",
  1344905L,    "Carboplatin",       "Cytotoxic",
  1378382L,    "Paclitaxel",        "Cytotoxic",
  1315942L,    "Docetaxel",         "Cytotoxic",
  1314924L,    "Gemcitabine",       "Cytotoxic",
  1305058L,    "Methotrexate",      "Cytotoxic",
  45775965L,   "Pembrolizumab",     "Immunotherapy",
  45892628L,   "Nivolumab",         "Immunotherapy",
  1387104L,    "Trastuzumab",       "Targeted",
  1397141L,    "Bevacizumab",       "Targeted",
  1304107L,    "Imatinib",          "Targeted",
  1314273L,    "Rituximab",         "Targeted",
  1436678L,    "Tamoxifen",         "Hormonal",
  1315946L,    "Letrozole",         "Hormonal"
)

drug_counts <- chemo_drugs %>%
  rowwise() %>%
  mutate(
    stats = list({
      descs <- cdm$concept_ancestor %>%
        filter(ancestor_concept_id == concept_id) %>%
        pull(descendant_concept_id)
      ids <- unique(c(concept_id, descs))
      cdm$drug_exposure %>%
        filter(drug_concept_id %in% ids) %>%
        summarise(persons = n_distinct(person_id), records = n()) %>%
        collect()
    })
  ) %>%
  ungroup() %>%
  tidyr::unnest(stats) %>%
  select(drug_name, drug_class, concept_id, persons, records) %>%
  arrange(desc(persons))
```

```{r chemo-drugs-table}
drug_counts %>%
  mutate(across(c(persons, records), fmt)) %>%
  knitr::kable(
    col.names = c("Drug", "Class", "Concept ID", "Persons", "Records"),
    align = c("l", "l", "r", "r", "r")
  )
```

```{r chemo-drugs-plot, fig.height=7}
drug_counts %>%
  filter(persons > 0) %>%
  mutate(drug_label = paste0(drug_name, " (", drug_class, ")")) %>%
  mutate(drug_label = reorder(drug_label, persons)) %>%
  ggplot(aes(x = persons, y = drug_label, fill = drug_class)) +
  geom_col() +
  geom_text(aes(label = fmt(persons)), hjust = -0.1, size = 3.2) +
  scale_fill_brewer(palette = "Set2", name = "Drug Class") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.25)),
                     labels = scales::comma) +
  labs(
    title = "Chemotherapy & Antineoplastic Drug Exposure by Person Count",
    x = "Persons", y = NULL
  )
```

## Drug class summary

```{r drug-class-summary}
drug_counts %>%
  group_by(drug_class) %>%
  summarise(
    agents = n(),
    agents_with_data = sum(persons > 0),
    total_persons = sum(persons),
    total_records = sum(records),
    .groups = "drop"
  ) %>%
  arrange(desc(total_persons)) %>%
  mutate(across(c(total_persons, total_records), fmt)) %>%
  knitr::kable(
    col.names = c("Drug Class", "Agents Checked", "Agents with Data",
                   "Total Persons", "Total Records"),
    align = c("l", "r", "r", "r", "r")
  )
```

# 4. Chemotherapy Administration Procedures

Beyond drug exposures, oncology data often records chemotherapy administration
as procedures. We search for a broad set of procedure concepts related to
chemotherapy delivery, radiation therapy, and oncologic surgery.

```{r onc-procedures}
onc_procedures <- tribble(
  ~concept_id, ~procedure_name,               ~category,
  4273629L,    "Chemotherapy",                 "Chemotherapy",
  4163713L,    "Chemotherapy cycle",           "Chemotherapy",
  4141762L,    "Oral chemotherapy",            "Chemotherapy",
  44790238L,   "Chemotherapy delivery",        "Chemotherapy",
  1242725L,    "Radiotherapy",                 "Radiation",
  4223087L,    "Radiation therapy mgmt",       "Radiation",
  4311405L,    "Biopsy",                       "Diagnostic",
  4146053L,    "Manual pelvic examination",    "Diagnostic",
  4306780L,    "Gynecologic examination",      "Diagnostic"
)

proc_counts <- onc_procedures %>%
  rowwise() %>%
  mutate(
    stats = list({
      descs <- cdm$concept_ancestor %>%
        filter(ancestor_concept_id == concept_id) %>%
        pull(descendant_concept_id)
      ids <- unique(c(concept_id, descs))
      cdm$procedure_occurrence %>%
        filter(procedure_concept_id %in% ids) %>%
        summarise(persons = n_distinct(person_id), records = n()) %>%
        collect()
    })
  ) %>%
  ungroup() %>%
  tidyr::unnest(stats) %>%
  select(procedure_name, category, concept_id, persons, records) %>%
  arrange(desc(persons))
```

```{r onc-procedures-table}
proc_counts %>%
  mutate(across(c(persons, records), fmt)) %>%
  knitr::kable(
    col.names = c("Procedure", "Category", "Concept ID", "Persons", "Records"),
    align = c("l", "l", "r", "r", "r")
  )
```

```{r onc-procedures-plot}
proc_counts %>%
  filter(persons > 0) %>%
  mutate(procedure_name = reorder(procedure_name, persons)) %>%
  ggplot(aes(x = persons, y = procedure_name, fill = category)) +
  geom_col() +
  geom_text(aes(label = fmt(persons)), hjust = -0.1, size = 3.2) +
  scale_fill_brewer(palette = "Set1", name = "Category") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.25)),
                     labels = scales::comma) +
  labs(
    title = "Oncology-Related Procedures by Person Count",
    x = "Persons", y = NULL
  )
```

# 5. Tumor Markers & Oncology Measurements

Tumor markers are key biomarkers in oncology research. We check for the most
commonly ordered lab tests: PSA, CA-125, CEA, and general tumor marker
measurements.

```{r tumor-markers}
tumor_markers <- tribble(
  ~concept_id, ~marker_name,
  3013603L,    "PSA (Prostate Specific Ag)",
  3037551L,    "CA-125 (Cancer Ag 125)",
  3003785L,    "CEA (Carcinoembryonic Ag)",
  3002131L,    "PSA [Units/vol]",
  4353605L,    "Tumor marker measurement"
)

marker_counts <- tumor_markers %>%
  rowwise() %>%
  mutate(
    stats = list({
      descs <- cdm$concept_ancestor %>%
        filter(ancestor_concept_id == concept_id) %>%
        pull(descendant_concept_id)
      ids <- unique(c(concept_id, descs))
      cdm$measurement %>%
        filter(measurement_concept_id %in% ids) %>%
        summarise(
          persons    = n_distinct(person_id),
          records    = n(),
          with_value = sum(!is.na(value_as_number))
        ) %>%
        collect()
    })
  ) %>%
  ungroup() %>%
  tidyr::unnest(stats) %>%
  select(marker_name, concept_id, persons, records, with_value) %>%
  arrange(desc(persons))
```

```{r tumor-markers-table}
marker_counts %>%
  mutate(
    value_pct = ifelse(records > 0, round(with_value / records * 100, 1), NA),
    across(c(persons, records, with_value), fmt)
  ) %>%
  knitr::kable(
    col.names = c("Marker", "Concept ID", "Persons", "Records",
                   "With Numeric Value", "Value %"),
    align = c("l", "r", "r", "r", "r", "r")
  )
```

# 6. Cancer Staging

Staging information is essential for oncology studies but is often sparse in
EHR-derived CDM data. We check for both SNOMED staging concepts and the
NAACCR-based TNM staging variables.

```{r staging}
staging_concepts <- tribble(
  ~concept_id, ~staging_name,                  ~source,
  4111627L,    "Tumor staging",                "SNOMED",
  607184L,     "Malignant tumor staging",      "SNOMED",
  4161178L,    "Tumor stage",                  "SNOMED",
  4121177L,    "TNM tumor staging",            "SNOMED",
  35918286L,   "TNM Path Stage Group",         "NAACCR",
  35918597L,   "TNM Clin Stage Group",         "NAACCR",
  4300140L,    "Tumor stage finding",          "SNOMED"
)

staging_meas <- staging_concepts %>%
  filter(source != "SNOMED" | !staging_name %in% c("Tumor stage finding")) %>%
  rowwise() %>%
  mutate(
    stats = list({
      descs <- cdm$concept_ancestor %>%
        filter(ancestor_concept_id == concept_id) %>%
        pull(descendant_concept_id)
      ids <- unique(c(concept_id, descs))
      cdm$measurement %>%
        filter(measurement_concept_id %in% ids) %>%
        summarise(persons = n_distinct(person_id), records = n()) %>%
        collect()
    })
  ) %>%
  ungroup() %>%
  tidyr::unnest(stats) %>%
  select(staging_name, source, concept_id, persons, records)

# Also check condition_occurrence for staging findings
staging_cond <- staging_concepts %>%
  filter(staging_name == "Tumor stage finding") %>%
  rowwise() %>%
  mutate(
    stats = list({
      descs <- cdm$concept_ancestor %>%
        filter(ancestor_concept_id == concept_id) %>%
        pull(descendant_concept_id)
      ids <- unique(c(concept_id, descs))
      cdm$condition_occurrence %>%
        filter(condition_concept_id %in% ids) %>%
        summarise(persons = n_distinct(person_id), records = n()) %>%
        collect()
    })
  ) %>%
  ungroup() %>%
  tidyr::unnest(stats) %>%
  mutate(source = "SNOMED (condition)") %>%
  select(staging_name, source, concept_id, persons, records)

staging_all <- bind_rows(staging_meas, staging_cond) %>%
  arrange(desc(persons))
```

```{r staging-table}
staging_all %>%
  mutate(across(c(persons, records), fmt)) %>%
  knitr::kable(
    col.names = c("Staging Concept", "Source", "Concept ID", "Persons", "Records"),
    align = c("l", "l", "r", "r", "r")
  )
```

```{r staging-note, results='asis', echo=FALSE}
has_staging <- any(staging_all$persons > 0)
if (has_staging) {
  cat("> **Staging data is present.** This dataset contains tumor staging",
      "information, which is valuable for stratifying analyses by disease",
      "severity.\n")
} else {
  cat("> **No staging data found.** This is common in EHR-derived CDM",
      "datasets where staging is documented in unstructured clinical notes",
      "rather than discrete fields. Consider supplementing with tumor",
      "registry data (e.g., NAACCR) if staging is required for your study.\n")
}
```

# 7. Disease Progression & Remission

Evidence of disease trajectory is important for longitudinal oncology studies.
We check for concepts representing cancer recurrence, progression, and
remission.

```{r progression}
trajectory_concepts <- tribble(
  ~concept_id, ~concept_label,                          ~domain,
  764225L,     "Recurrent malignant neoplastic disease", "Condition",
  4168352L,    "Tumor progression",                      "Condition",
  432851L,     "Metastatic malignant neoplasm",          "Condition",
  37172812L,   "Malignant neoplasm in full remission",   "Condition",
  37172814L,   "Malignant neoplasm in partial remission", "Condition"
)

traj_counts <- trajectory_concepts %>%
  rowwise() %>%
  mutate(
    stats = list({
      descs <- cdm$concept_ancestor %>%
        filter(ancestor_concept_id == concept_id) %>%
        pull(descendant_concept_id)
      ids <- unique(c(concept_id, descs))
      cdm$condition_occurrence %>%
        filter(condition_concept_id %in% ids) %>%
        summarise(persons = n_distinct(person_id), records = n()) %>%
        collect()
    })
  ) %>%
  ungroup() %>%
  tidyr::unnest(stats) %>%
  select(concept_label, concept_id, persons, records) %>%
  arrange(desc(persons))
```

```{r progression-table}
traj_counts %>%
  mutate(across(c(persons, records), fmt)) %>%
  knitr::kable(
    col.names = c("Concept", "Concept ID", "Persons", "Records"),
    align = c("l", "r", "r", "r")
  )
```

```{r progression-plot}
traj_counts %>%
  filter(persons > 0) %>%
  mutate(concept_label = reorder(concept_label, persons)) %>%
  ggplot(aes(x = persons, y = concept_label)) +
  geom_col(fill = "#756bb1") +
  geom_text(aes(label = fmt(persons)), hjust = -0.1, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.25)),
                     labels = scales::comma) +
  labs(
    title = "Disease Trajectory Concepts",
    x = "Persons", y = NULL
  )
```

# 8. Summary Scorecard

This scorecard summarizes the oncology readiness of the dataset at a glance.

```{r scorecard}
scorecard <- tribble(
  ~dimension,                ~metric,
  "Total persons in CDM",    total_persons,
  "Persons with any cancer", cancer_summary$cancer_persons,
  "Cancer sites with data",  sum(site_counts$persons > 0),
  "Chemo drugs with data",   sum(drug_counts$persons > 0),
  "Onc procedures with data",sum(proc_counts$persons > 0),
  "Tumor markers with data", sum(marker_counts$persons > 0),
  "Staging concepts found",  sum(staging_all$persons > 0),
  "Trajectory concepts found", sum(traj_counts$persons > 0)
)

scorecard %>%
  mutate(metric = fmt(metric)) %>%
  knitr::kable(col.names = c("Dimension", "Count"), align = c("l", "r"))
```

```{r scorecard-assessment, results='asis', echo=FALSE}
n_cancer <- cancer_summary$cancer_persons
n_drugs  <- sum(drug_counts$persons > 0)
n_stage  <- sum(staging_all$persons > 0)

cat("## Assessment\n\n")

if (n_cancer == 0) {
  cat("This dataset contains **no identifiable cancer conditions**. It is",
      "not suitable for oncology research without supplemental data.\n")
} else if (n_drugs == 0 && n_stage == 0) {
  cat("This dataset contains **", fmt(n_cancer), " cancer patients** but",
      "**lacks chemotherapy drug exposures and staging data**. It may support",
      "descriptive epidemiology but is limited for treatment or outcomes",
      "studies.\n")
} else if (n_stage == 0) {
  cat("This dataset contains **", fmt(n_cancer), " cancer patients** with",
      "some treatment data but **no structured staging**. It may support",
      "treatment pattern analyses but will be limited for stage-stratified",
      "outcomes research.\n")
} else {
  cat("This dataset contains **", fmt(n_cancer), " cancer patients** with",
      "treatment data and staging information. It has strong potential for",
      "oncology research including treatment patterns and outcomes studies.\n")
}
```

# Disconnect

```{r disconnect}
cdmDisconnect(cdm)
```

---

*Report generated `r Sys.time()`*
