---
title: "run_etl: Detailed Steps and Mechanics"
author: "ETLDelphi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{run_etl: Detailed Steps and Mechanics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteKeyword{run_etl}
  %\VignetteKeyword{ETL}
  %\VignetteKeyword{OMOP}
  %\VignetteKeyword{DuckDB}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

`run_etl()` is the main entry point for the ETLDelphi package. It orchestrates the entire transformation from Delphi/Delfi source data into OMOP CDM v5.4 tables in DuckDB. This vignette describes **in detail** how `run_etl` works, what happens at each step, and the mechanics of the SQL execution pipeline.

---

## What run_etl Does

When you call `run_etl(con)`:

1. **Loads configuration** — Reads YAML from `inst/extdata/config.yml` (or a path you provide) to get schema names (`src`, `stg`, `cdm`) and type concept IDs.
2. **Locates SQL scripts** — Uses `inst/sql/` (or your custom `sql_dir`) as the script directory.
3. **Delegates to `run_sql_scripts()`** — Executes all `.sql` files in lexicographic order, substituting schema placeholders and logging each step.

The R package does **not** perform any data transformation itself. All business logic lives in SQL; the package only orchestrates execution.

---

## Internal Flow

### run_etl() Implementation

```r
run_etl <- function(con,
                    config_path = NULL,
                    config = NULL,
                    sql_dir = NULL,
                    dry_run = FALSE,
                    from_step = NULL,
                    to_step = NULL)
```

1. **Config resolution**: If `config` is `NULL`, load from `config_path` or default `system.file("extdata", "config.yml", package = "ETLDelphi")`. If `config` is provided, it overrides `config_path`.
2. **SQL directory resolution**: If `sql_dir` is `NULL`, use `system.file("sql", package = "ETLDelphi")`. Otherwise use the provided path.
3. **Call `run_sql_scripts()`** with `con`, `sql_dir`, `config`, `dry_run`, `from_step`, `to_step`.

### run_sql_scripts() Mechanics

`run_sql_scripts()` in `R/sql_runner.R` performs the actual work:

| Step | Action |
|------|--------|
| **1. Discover files** | `fs::dir_ls(sql_dir, recurse = TRUE, glob = "*.sql")` then `sort()` for lexicographic order. |
| **2. Filter by from_step / to_step** | If `from_step` is set, drop files before the first match; if `to_step` is set, drop files after the last match. Matching is by folder or file name prefix (e.g. `"40_cdm_core"`). |
| **3. Dry run** | If `dry_run = TRUE`, list files and return; do not execute. |
| **4. Assign run_id** | Use `as.integer(as.numeric(Sys.time()))` if not provided. |
| **5. Insert run log** | If `stg.etl_run_log` exists, insert a row with `run_id`, `started_at`, `status = 'running'`. |
| **6. Loop over files** | For each `.sql` file: |
| 6a | Read file contents with `readLines()`. |
| 6b | **Schema substitution**: Replace `@cdmDatabaseSchema`, `{cdm}`, `{stg}`, `{src}` with values from `config$schemas`. |
| 6c | Strip single-line SQL comments (`-- ...`) so semicolons in comments do not break statement splitting. |
| 6d | Split into statements via `split_sql_statements()` (semicolon-terminated, respecting quoted strings). |
| 6e | Execute each non-empty statement with `DBI::dbExecute(con, stmt)`. |
| 6f | On error, abort with `cli::cli_abort()` and report the step name. |
| 6g | If `stg.etl_step_log` exists, insert a row with `run_id`, `step_name`, `file_name`, `started_at`, `ended_at`, `status`, `rows_affected`, `notes`. |
| **7. Mark run complete** | Update `stg.etl_run_log` with `ended_at` and `status = 'completed'`. |

---

## Execution Order: The Seven Phases

Scripts are run in strict order. The numbered folders and files ensure correct dependencies (e.g. staging before keys, keys before vocab, vocab before CDM core, etc.).

### Phase 0 — Admin

| File | Purpose |
|------|---------|
| `00_admin/00_create_stg_schema.sql` | `CREATE SCHEMA IF NOT EXISTS stg`. Creates the staging schema; the ETL writes all intermediate tables here. |
| `00_admin/01_create_etl_log_tables.sql` | Creates `stg.etl_run_log` and `stg.etl_step_log` for run and step metadata. These tables are used by `run_sql_scripts` to log progress. |
| `00_admin/02_clear_stg_tables.sql` | Drops all `stg.*` tables (reject, map, staged, QC) in dependency order. Does **not** touch `src.*` or `cdm.*`. Allows the ETL to be re-run from a clean state. |

### Phase 1 — Staging (10_stage)

Source tables in `src` have varchar columns. Each staging script:

1. Parses dates with `try_strptime()` (multiple formats)
2. Casts numerics with `try_cast()`
3. Trims and normalizes strings (e.g. `LOWER(TRIM(...))`)
4. Writes rows that **fail** parsing to a corresponding `stg.reject_*` table

| Script | Output table | Reject table |
|--------|--------------|--------------|
| `10_stage_enrollment.sql` | `stg.enrollment` | `stg.reject_enrollment_dates` |
| `11_stage_provider.sql` | `stg.provider` | `stg.reject_provider` |
| `12_stage_encounter.sql` | `stg.encounter` | `stg.reject_encounter` |
| `13_stage_death.sql` | `stg.death` | `stg.reject_death` |
| `14_stage_problem.sql` | `stg.problem` | `stg.reject_problem` |
| `15_stage_med_orders.sql` | `stg.medication_orders` | `stg.reject_med_orders` |
| `16_stage_med_fulfillment.sql` | `stg.medication_fulfillment` | `stg.reject_med_fulfillment` |
| `17_stage_current_meds.sql` | `stg.current_medications` | `stg.reject_current_meds` |
| `18_stage_immunization.sql` | `stg.immunization` | `stg.reject_immunization` |
| `19_stage_lab_orders.sql` | `stg.lab_orders` | `stg.reject_lab_orders` |
| `20_stage_lab_results.sql` | `stg.lab_results` | `stg.reject_lab_results` |
| `21_stage_vital_sign.sql` | `stg.vital_sign` | `stg.reject_vital_sign` |
| `22_stage_allergy.sql` | `stg.allergy` | `stg.reject_allergy` |
| `23_stage_therapy_orders.sql` | `stg.therapy_orders` | — |
| `24_stage_therapy_actions.sql` | `stg.therapy_actions` | — |

Example: `10_stage_enrollment.sql` parses `DOB` with `COALESCE(try_strptime(...), ...)`; rows where DOB is non-null but unparseable go to `stg.reject_enrollment_dates`.

### Phase 2 — Keys (20_keys)

Deterministic integer IDs for OMOP entities. Each map uses `ROW_NUMBER() OVER (ORDER BY <stable_key>)` so the same source data yields the same IDs on every run.

| Script | Map table | Source of keys |
|--------|-----------|----------------|
| `30_map_person.sql` | `stg.map_person(member_id, person_id)` | Distinct `member_id` from union of enrollment, encounter, death, problem, medication_orders, immunization, lab_results, vital_sign, allergy, therapy_orders, therapy_actions, current_medications |
| `31_map_provider.sql` | `stg.map_provider(provider_id_source, provider_id)` | Distinct Provider_ID from provider, encounter, medication_orders, immunization, lab_results, problem |
| `32_map_visit.sql` | `stg.map_visit(encounter_id_source, visit_occurrence_id)` | Distinct Encounter_ID from `stg.encounter` |
| `33_map_location.sql` | `stg.map_location(location_key, location_id)` | Distinct concatenation of address_1, address_2, city, state, zip from `stg.enrollment` |
| `34_map_care_site.sql` | `stg.map_care_site(care_site_key, care_site_id)` | Distinct Clinic_ID from `stg.encounter` |

### Phase 3 — Vocabulary mapping (30_vocab)

Concept maps resolve source values to OMOP concept IDs. Unmapped values → `0` (per `concept_id_unknown` in config).

| Script | Map table | Strategy |
|--------|-----------|----------|
| `40_map_gender.sql` | `stg.map_gender` | Hardcoded: female/f→8532, male/m→8507, unknown/other→8551; enrollment-specific values inserted with 0 |
| `41_map_race.sql` | `stg.map_race` | Lookup: white, black, asian, etc. to standard race concept_ids |
| `42_map_visit_concept.sql` | `stg.map_visit_concept` | (appt_type, clinic_type) → visit_concept_id; default 44813942 (Outpatient) |
| `43_map_drug.sql` | `stg.map_drug_order` | NDC → `cdm.concept` (vocabulary_id='NDC') → concept_relationship 'Maps to' → RxNorm |
| `44_map_loinc_measurement.sql` | `stg.map_loinc_measurement` | Test_LOINC → `cdm.concept` (vocabulary_id='LOINC') |
| `45_map_units.sql` | `stg.map_units` | Unit strings (mg, cm, kg, etc.) → unit_concept_id |
| `46_map_condition.sql` | `stg.map_condition` | Problem_Code → concept (ICD10/ICD9/SNOMED/Medcin) → Maps to standard |
| `47_map_vitals.sql` | `stg.map_vitals` | Hardcoded: vital name → measurement_concept_id and default unit |
| `48_map_allergy.sql` | `stg.map_allergy` | v1: observation_concept_id = 0; source_value preserved |
| `49_map_therapy.sql` | `stg.map_therapy` | Code + Vocabulary → concept (CPT4, HCPCS, SNOMED, ICD10PCS) → Maps to standard |

### Phase 4 — CDM core (40_cdm_core)

Load order respects foreign keys: location and care_site first, then provider, person, visit, note, death.

| Script | CDM table | Source |
|--------|-----------|--------|
| `50_load_location.sql` | `cdm.location` | `stg.map_location` + enrollment address fields |
| `51_load_care_site.sql` | `cdm.care_site` | `stg.map_care_site` |
| `52_load_provider.sql` | `cdm.provider` | `stg.map_provider` + `stg.provider` |
| `53_load_person.sql` | `cdm.person` | `stg.map_person` + `stg.enrollment` (only rows with parseable DOB) |
| `54_load_visit_occurrence.sql` | `cdm.visit_occurrence` | `stg.encounter` + maps |
| `55_load_note.sql` | `cdm.note` | SOAP_Note from encounter |
| `56_load_death.sql` | `cdm.death` | `stg.death` + `stg.map_person` |
| `57_load_cdm_source.sql` | `cdm.cdm_source` | One row from vignette metadata: name, abbreviation, holder, description, ETL reference, release dates, CDM version, vocabulary version |

Person load: only inserts when `year_of_birth` can be derived from `stg.enrollment.dob_date`. Members without enrollment or parseable DOB are logged in `stg.reject_person_missing_dob`.

### Phase 5 — CDM clinical (50_cdm_clinical)

| Script | CDM table | Source |
|--------|-----------|--------|
| `60_load_condition_occurrence.sql` | `cdm.condition_occurrence` | `stg.problem` + `stg.map_condition` |
| `61_load_drug_exposure_orders.sql` | `cdm.drug_exposure` | `stg.medication_orders` (drug_type 38000177) |
| `62_load_drug_exposure_fulfillment.sql` | `cdm.drug_exposure` | `stg.medication_fulfillment` (drug_type 38000230) |
| `63_load_drug_exposure_immunization.sql` | `cdm.drug_exposure` | `stg.immunization` (drug_type 38000280) |
| `64_load_measurement_vitals.sql` | `cdm.measurement` | `stg.vital_sign` unpivoted |
| `65_load_measurement_labs.sql` | `cdm.measurement` | `stg.lab_results` |
| `66_load_observation_allergy.sql` | `cdm.observation` | `stg.allergy` |
| `67_load_procedure_therapy.sql` | `cdm.procedure_occurrence` | `stg.therapy_orders` + `stg.therapy_actions` |

### Phase 6 — CDM derived (60_cdm_derived)

| Script | CDM table | Logic |
|--------|-----------|--------|
| `80_load_observation_period.sql` | `cdm.observation_period` | Per person: min/max of all event dates from visit, condition, drug, measurement, observation, procedure, death |

### Phase 7 — QC (90_qc)

| Script | QC table | Content |
|--------|----------|---------|
| `90_row_counts.sql` | `stg.qc_row_counts` | Row counts for key stg.* and cdm.* tables |
| `91_mapping_coverage.sql` | `stg.qc_mapping_coverage` | Per clinical table: total rows, unmapped count, unmapped % |
| `92_referential_integrity.sql` | `stg.qc_orphans` | Orphan checks (person_id, visit_occurrence_id) |
| `93_parse_failures.sql` | `stg.qc_parse_failures` | Counts per reject table |
| `94_required_fields.sql` | `stg.qc_required_fields` | NOT NULL violations |

---

## Schema Substitution

SQL scripts use placeholders that are replaced at runtime:

| Placeholder | Replaced with |
|-------------|----------------|
| `@cdmDatabaseSchema` | `config$schemas$cdm` (default: `cdm`) |
| `{cdm}` | `config$schemas$cdm` |
| `{stg}` | `config$schemas$stg` |
| `{src}` | `config$schemas$src` |

This allows the same scripts to run against different schema names without editing SQL.

---

## Using run_etl

### Full run

```{r full-run}
library(DBI)
library(duckdb)
library(ETLDelphi)

con <- dbConnect(duckdb::duckdb(), "path/to/your.duckdb")
run_etl(con)
dbDisconnect(con, shutdown = TRUE)
```

### Dry run (list scripts only)

```{r dry-run}
run_etl(con, dry_run = TRUE)
```

### Subset of steps

```{r from-to}
# From admin through clear (does not run staging)
run_etl(con, from_step = "00_admin", to_step = "02_clear")

# From CDM core through end (assumes stg already populated)
run_etl(con, from_step = "40_cdm_core")

# Only QC
run_etl(con, from_step = "90_qc", to_step = "94_required_fields")
```

### Custom config or SQL directory

```{r custom}
run_etl(con, config_path = "/path/to/config.yml")
run_etl(con, config = list(schemas = list(src = "src", stg = "stg", cdm = "cdm")))
run_etl(con, sql_dir = "/path/to/sql")
```

---

## Error Handling

- **First failure stops the run**: If any SQL statement throws an error, `run_sql_scripts` calls `cli::cli_abort()` and reports the step name (e.g. `10_stage/10_stage_enrollment.sql`).
- **Logging**: Before aborting, the step is logged to `stg.etl_step_log` with `status = 'error'` and `notes` containing the error message.
- **Run status**: The run remains `status = 'running'` in `stg.etl_run_log` when aborted (no automatic rollback to `completed`).

---

## Logging Tables

| Table | Purpose |
|-------|---------|
| `stg.etl_run_log` | One row per run: `run_id`, `started_at`, `ended_at`, `status`, `git_sha` |
| `stg.etl_step_log` | One row per script: `run_id`, `step_name`, `file_name`, `started_at`, `ended_at`, `status`, `rows_affected`, `notes` |

These tables are created by `01_create_etl_log_tables.sql` and populated by `run_sql_scripts` during execution.

---

## Summary

`run_etl` is a thin orchestration layer that:

1. Loads config and resolves paths
2. Calls `run_sql_scripts`, which discovers `.sql` files in lexicographic order
3. Substitutes schema placeholders and executes each script
4. Logs each step and the overall run
5. Aborts on the first error

All transformation logic lives in 54 SQL scripts across 7 phases: admin → staging → keys → vocab → cdm_core → cdm_clinical → cdm_derived → qc. The pipeline is deterministic, idempotent (with `02_clear_stg_tables`), and observable via reject tables and QC tables.
