---
title: "run_etl: Detailed Steps and Mechanics"
author: "ETLDelphi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{run_etl: Detailed Steps and Mechanics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteKeyword{run_etl}
  %\VignetteKeyword{ETL}
  %\VignetteKeyword{OMOP}
  %\VignetteKeyword{DuckDB}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

`run_etl()` is the main entry point for the core Delphi-to-OMOP pipeline.

Its job is narrow:

- resolve configuration
- locate the SQL scripts
- call `run_sql_scripts()`

All data transformation rules live in SQL. The R layer is the execution engine.

## What `run_etl()` Actually Does

The public function signature is:

```r
run_etl(
  con,
  config_path = NULL,
  config = NULL,
  sql_dir = NULL,
  dry_run = FALSE,
  from_step = NULL,
  to_step = NULL
)
```

The control flow is:

1. If `config` is supplied, use it.
2. Otherwise, if `config_path` is supplied, read YAML from that path.
3. Otherwise, call `default_etl_config()`.
4. Resolve `sql_dir` to `system.file("sql", package = "ETLDelphi")` unless you override it.
5. Call `run_sql_scripts()` with the resolved arguments.

That means the default behavior is driven by `default_etl_config()`, not by a bundled YAML file.

## How `run_sql_scripts()` Executes The Pipeline

`run_sql_scripts()` in `R/sql_runner.R` performs the real work.

### 1. Discover SQL Files

It recursively finds all `.sql` files under `sql_dir` and sorts them lexicographically.

That ordering is the pipeline contract. Folder prefixes such as `00_admin`, `10_stage`, and `90_qc` are what make the run deterministic.

### 2. Apply `from_step` / `to_step`

If you pass `from_step` or `to_step`, the runner filters the file list before any SQL executes.

Matching works on:

- the relative path, such as `40_cdm_core/54_load_visit_occurrence.sql`
- the file basename prefix, such as `54_load_visit_occurrence`

### 3. Support Dry Runs

With `dry_run = TRUE`, the runner prints the filtered file list and returns without executing any statements.

This is useful when you need to confirm exactly which scripts will run after a refactor.

### 4. Initialize Logging

The runner creates a numeric `run_id` if one is not provided and attempts to insert a row into `stg.etl_run_log`.

This insert is conditional: it only happens if the log table already exists.

That is why `00_admin/01_create_etl_log_tables.sql` needs to run before you expect full logging coverage.

### 5. Read And Rewrite Each SQL File

For each script, the runner:

- reads the file contents
- substitutes schema placeholders
- substitutes type concept placeholders
- injects resolved file paths for mapping CSVs
- strips single-line SQL comments
- splits the script into statements with `split_sql_statements()`

The most important substitutions are:

| Placeholder / pattern | Replacement |
|-----------------------|-------------|
| `@cdmDatabaseSchema` | configured CDM schema |
| `{cdm}` | configured CDM schema |
| `{stg}` | configured staging schema |
| `{src}` | configured source schema |
| literal `cdm.` | configured CDM schema plus `.` |
| `{visit_type_concept_id}` and similar | values from `config$type_concept_ids` |
| `@drugNameMappingPath` | resolved drug-name CSV path |
| `@customMappingPath` | resolved custom concept mapping CSV path |
| `@customNdcMappingPath` | resolved custom NDC mapping CSV path |

The literal `cdm.` rewrite matters because many SQL scripts reference `cdm.concept` or other CDM tables directly. That allows the same SQL to run when the CDM schema is not named `main`.

### 6. Execute Statement By Statement

After splitting, the runner executes each non-empty statement with `DBI::dbExecute()`.

This avoids relying on multi-statement execution behavior and makes it easier to identify the failing step.

### 7. Write Step Logs

After each file completes, the runner attempts to insert a row into `stg.etl_step_log` with:

- `run_id`
- `step_name`
- `file_name`
- start and end timestamps
- status
- the last `rows_affected` value returned by `dbExecute()`
- optional notes

Again, this insert is conditional on the table existing.

### 8. Mark The Run Completed

If `stg.etl_run_log` exists, the runner updates that row to `status = 'completed'` at the end.

## Error Behavior

The runner stops on the first SQL error.

When a statement fails:

- the current file is reported with `cli::cli_abort()`
- the step loop stops immediately
- previously completed steps remain committed in the DuckDB database

The ETL does not wrap the whole run in a global transaction.

That is intentional. ETLDelphi is built to be inspectable after failure rather than opaque behind a rollback.

## Current Phase Order

The current release runs the core ETL in this folder order:

1. `00_admin`
2. `10_stage`
3. `20_keys`
4. `30_vocab`
5. `40_cdm_core`
6. `45_nlp`
7. `50_cdm_clinical`
8. `60_cdm_derived`
9. `90_qc`

The exact file count can change between releases, but the ordering rule does not.

## Practical Examples

### Full Run

```{r full-run}
con <- DBI::dbConnect(duckdb::duckdb(), "path/to/delphi.duckdb")
run_etl(con)
DBI::dbDisconnect(con, shutdown = TRUE)
```

### Preview The Exact Files

```{r preview}
run_etl(con, dry_run = TRUE)
```

### Re-run Only The Mapping And Downstream Phases

```{r rerun}
run_etl(con, from_step = "30_vocab")
```

### Run Only The QC Phase

```{r qc-only}
run_etl(con, from_step = "90_qc")
```

### Use A Modified Config

```{r config}
cfg <- default_etl_config()
cfg$schemas$cdm <- "omop"

run_etl(con, config = cfg)
```

## Why This Design Is Useful

This split between R orchestration and SQL transforms gives you:

- a single place to inspect the run order (`run_sql_scripts()`)
- a stable path to override concepts and mapping files without editing SQL
- the ability to rerun from any phase for debugging
- deterministic, reproducible script execution

It also keeps SQL review straightforward. If a data transformation is wrong, the fix should usually be in `inst/sql`, not in `run_etl()`.

## Summary

`run_etl()` is intentionally small. The mechanics that matter are all in `run_sql_scripts()`:

1. discover files
2. filter the run window
3. substitute schemas, concepts, and file paths
4. split and execute statements
5. log what happened
6. stop at the first failure

That makes the ETL easy to reason about and easy to review when something changes.
